\bf Model Visualization Techniques for a Social Network Model

### 2018-07-11 paper-child.Rnw
ugh get this done

<<pkgs>>=
# if (packageVersion("ggplot2") > "2.2.0"){
#   remove.packages("ggplot2")
#   install.packages("ggplot2")
# }
library(ggplot2)
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
#devtools::install_github("sctyner/netvizinf")
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

@@

\section{Introduction}
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks.
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
%XXX - okay, here is what i have done for that:
Social networks, such as collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations have been studied for decades.  
%beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of . 
In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. %There are many kinds of social networks, but there are not as many statistical models for social network data. 
Some network models that have been applied to social networks include the exponential random graph model (ERGM) and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they are not able to model network change as time passes. When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time. 
Models for dynamic social networks have a great deal of potential because of how realistic they can be. A social network does not form spontaneously: it \textit{evolves} over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par We concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this,
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user. In order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out into the light. %, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By visualizing the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

\par To get behind the scenes of the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output. Stochastic actor-oriented models are an excellent example of a family of models that can benefit greatly from the application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

Network data is of frequent interest to researchers in a wide array of fields, such as biologists studying neural networks and sociologists studying social networks.% There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}. 
% Each of these examples have one thing in common:
Network data across fields have a similar data structure. There are always units of observation, such as neurons %the power stations, websites, neurons, 
and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}. There are also always connections of some kind between those units, such as electrical signals and relationships, which we will call \textit{edges} or \textit{ties}. Networks often change over time, such as when there are relationships formed in a friendship network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a friendship network and the length of a friendship.  %the institution of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures pose unique problems to analysts. Some questions that researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network? Do differences in node variables affect the formation or dissolution of edges? Which views of the data are most informative for communicating significant model effects? What about other results of statistical analyses of the network? %These are, of course, just a few broad questions, and we focus here on the latter, which 
We aim to answer some of these questions through visual exploration of network data and models.

@@

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations. %If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct. If the variables are categorical, bar charts and mosaic plots are available to to the researcher. If the data are spatial, there is a well-defined region in which to view information. 
Network data, however, are much less cut-and-dried.

@@

There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines when there is an edge between the two nodes. These lines can also have arrows on them indicating the direction of the edge for directed networks. But because there is typically no natural placement of the points unless they have inherent spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}. Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
%hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
We only consider the presence or absence of an edge, but if the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix can be the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$ for all $i \neq j$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks if and only if every edge between two nodes is reciprocated.
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

Each type of visualization comes with its own advantages and disadvantages. For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges. Adjacency matrix visualization can be useful when the network is very complex, dense, or large. Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.  %\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.}
This is advantageous for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, however there is not one "correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.

\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}

% \st{What do I need to communicate in this section?
% \begin{itemize}
% \item What the models are for: dynamic social networks with actor-level covariates
% \item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit?
% \end{itemize}}

%The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes. The object of analysis is a social network that has been observed at several discrete time points. The nodes in the network can have associated covariate variables that may be incorporated into modeling the tie changes in the network. The model itself has two primary pieces: the rate of change of ties, and the objective function of the nodes, which determines the exact change that gets made.

%We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$. The first observation, $x(t_1)$ is always conditioned on. The number of nodes / actors / vertices is constant in all time points and is denoted $n$.  The tie variables are directed and  binary, and are denoted at $x_{ij}(t_m)$ for time point $m$, and this is sometimes abbreviated to $x_{ij}$ in notation. If there is a tie from node $i$ to node $j$, then $x_{ij}=1$, and $x_{ij}=0$ otherwise.

%\hh{mostly lit review plus a detailed description of the model and fitting process.}

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: edge, node, and temporal information. It models the change of a network over time, allowing for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways. Most actors in social networks have inherent properties that could affect how they change their role within the network, and ties between actors may affect formation or dissolution of ties.

@@

We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process, we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

@@

In the broadest definition of a SAOM, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, and thus our rate function is just a set of rate parameters in the overall model.  %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
%Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function,
In the simple model with a rate parameters instead of a rate function, the rate parameters dictate how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ If $j = i$, no change in the network is made. We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. %Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. 
For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence of nodes in the network given their current ties is expressed in Equation~\ref{eq:c1}.

@@

 The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$. %The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
%\begin{equation} \label{eq:c2}
% \tau(x|m) \sim Exp(n\alpha_m)
%\end{equation}
The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter is straightforward: the method of moments is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. %In order to achieve the memorylessness property of a Markov process,

% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
% \st{XXX see line 1257 for plot. where should I put it? }
% \hh{XXX Put a reference to the plot here.}

@@

%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time. The node $i$, which is the node that is changing at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC. For the ego node, $i$, in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. %\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?}
Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created that is currently $x_{ij}=0$, or no change will occur.

>>>>

<<pkgs>>=
# if (packageVersion("ggplot2") > "2.2.0"){
#   remove.packages("ggplot2")
#   install.packages("ggplot2")
# }
library(ggplot2)
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
# devtools::install_github("sctyner/netvizinf")
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

@@

\section{Introduction}

Social networks, such as collaboration networks between academic researchers or friendship networks in a school or university have been studied for decades. Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure. Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes. When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time. Models for dynamic social networks hold a great deal of potential because of how realistic they can be. Social networks do not form spontaneously, but rather they \textit{evolve} over time as edges are formed and dissolved, and new nodes join the social structure. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par To model dynamic social networks, we use stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are different from other network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network. For example, we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this information on shared interests in the modeling process.

\par Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models. SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the many simulations involved in parameter estimation are largely hidden from the software user. In order to better understand the model-fitting process of \texttt{RSiena}, we attempt to bring model fits and the fitting process to the foreground by visualizing the underlying methods and structures. Utimately, we aim to help researchers working with the models better understand the structure of and results from these models.

\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and explore the algorithms fitting the SAOMs instead of studying only the final output. SAOMs are an excellent example of a family of models that can reap many benefits from the application of model visualization techniques. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Visualizing the CTMC can provide researchers with more insight into the underlying features of the model. In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model. These simulations are hidden and unsaved by default for computational efficiency. With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

Network data across fields have similar data structures. There are always units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people. Networks often change over time, like when new relationships are formed in a social network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.).

These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.

@@

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations. Network data, however, carry no such inherent structure.

@@

There are two primary methods used to visualize networks that are very different from most traditional data visualizations. They are the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$. This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}. Because there is usually no natural placement of the points (the exception is nodes with spatial locations), a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.

\par Some commonly used layout algorithms, such as  the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.

\par The second method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\end{equation}
We only consider the presence or absence of an edge, but if the network has weighted edges, the entries in the adjacency matrix are the edge weights. An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row and column representing a node. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.

\par Each visualization comes with its own advantages and disadvantages. For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges. Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization. This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. 
Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.

\section{Defining SAOMs} \label{sec:saoms}

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information. It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways. Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.

@@

We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

@@

In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Then, we express this in Equation~\ref{eq:c1}.

@@

From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.

@@

Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.

According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. (These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model. The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.

The $\boldsymbol{\beta}$ parameters are either structural or covariate parameters. The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are only functions of the current network state, $x$. They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$. A table of some possible structural and covariate parameters and their corresponding statistics is given in Table~\ref{tab:effects}.
